Finiteness in Cubical Type Theory

### 2020-07-03 sections/search.tex
put back in old proof automation section

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
practical applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Proof Automation And Search Techniques}
For this prover we will not resort to reflection or similar techniques: instead,
we will trick the type checker to do our automation for us.
This is a relatively common technique, although not so much outside of Agda, so
we will briefly explain it here.

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, we need to be able to ``lift'' decidability back over a
function arrow.
For instance, given \(x\), \(y\), and \(z\) we already have
\(\AgdaDatatype{Dec}\;((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\)
(because equality over booleans is decidable).
In order to turn this into a proof that \AgdaFunction{∧} is associative we need
\(\AgdaDatatype{Dec}\;(\forall \; x \; y \; z \rightarrow (x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\).
The ability to do this is described formally by the notion of
``Exhaustibility''.
\begin{agdalisting}
  \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{exhaustible}
\end{agdalisting}
We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
\(A\), the universal quantification of the predicate is decidable.

@@

The proof that this type is finite mirrors its definition closely:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{expr-finite}
\end{agdalisting}
\subsection{Filtering to Correct Expressions}
We now have a way to construct, formally, every expression we can generate from
a given list of numbers.
This is incomplete in two ways, however.
Firstly, some expressions are invalid: we should not, for instance, be able to
divide two numbers which do not divide evenly.
Secondly, we are only interested in those expressions which actually represent
solutions: those which evaluate to the target, in other words.
We can write a function which tells us if both of these things hold for a given
expression like so:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Countdown.tex]{eval}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Countdown.tex]{app-op}
  \end{agdalisting}
\end{multicols}

With this all together, we can finally write down the type of all solutions to a
given countdown problem.
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{solution-type}
\end{agdalisting}
And, because the predicate here is decidable and a mere proposition, we can
prove that there are finitely many solutions:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{finite-solution}
\end{agdalisting}
And we can apply this to a particular problem like so:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{example-solution}
\end{agdalisting}

\section{Automating Proofs}

>>>>

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
practical applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, we need to be able to ``lift'' decidability back over a
function arrow.
For instance, given \(x\), \(y\), and \(z\) we already have
\(\AgdaDatatype{Dec}\;((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\)
(because equality over booleans is decidable).
In order to turn this into a proof that \AgdaFunction{∧} is associative we need
\(\AgdaDatatype{Dec}\;(\forall \; x \; y \; z \rightarrow (x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\).
The ability to do this is described formally by the notion of
``Exhaustibility''.
\begin{agdalisting}
  \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{exhaustible}
\end{agdalisting}
We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
\(A\), the universal quantification of the predicate is decidable.

@@

The proof that this type is finite mirrors its definition closely:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{expr-finite}
\end{agdalisting}
\subsection{Filtering to Correct Expressions}
We now have a way to construct, formally, every expression we can generate from
a given list of numbers.
This is incomplete in two ways, however.
Firstly, some expressions are invalid: we should not, for instance, be able to
divide two numbers which do not divide evenly.
Secondly, we are only interested in those expressions which actually represent
solutions: those which evaluate to the target, in other words.
We can write a function which tells us if both of these things hold for a given
expression like so:

\begin{minipage}{\linewidth}
  \begin{multicols}{2}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Countdown.tex]{eval}
    \end{agdalisting} \columnbreak
    \begin{agdalisting}
      \ExecuteMetaData[agda/Countdown.tex]{app-op}
    \end{agdalisting}
  \end{multicols}
\end{minipage}

With this all together, we can finally write down the type of all solutions to a
given countdown problem.
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{solution-type}
\end{agdalisting}
And, because the predicate here is decidable and a mere proposition, we can
prove that there are finitely many solutions:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{finite-solution}
\end{agdalisting}
And we can apply this to a particular problem like so:
\begin{agdalisting}
  \ExecuteMetaData[agda/Countdown.tex]{example-solution}
\end{agdalisting}
\section{Proof Automation}
So we have shown now how powerful the library is for proof search: it can
search for functions, subsets of types, products, and so on, completely
verifiable.
In this section we will present the more user-friendly interface to the library,
designed to be used to automate away tedious proofs in an easy way.
\subsection{How to make the Typechecker do Automation}
For this prover we will not resort to reflection or similar techniques: instead,
we will trick the type checker to do our automation for us.
This is a relatively common technique, although not so much outside of Agda, so
we will briefly explain it here.

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\subsection{Building an Interface}
We will next look at a syntax and interface to this proof technique.
It starts with the following two functions:

\begin{minipage}{\linewidth}
  \begin{multicols}{2}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Search.tex]{forall-ask}
    \end{agdalisting} \columnbreak
    \begin{agdalisting}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Search.tex]{exists-ask}
    \end{agdalisting}
  \end{multicols}
\end{minipage}
Clearly they're just restatements of exhaustibility and omniscience.
However, we can combine these functions with what we've constructed above to
create an automation procedure for each:

\begin{minipage}{\linewidth}
  \begin{multicols}{2}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Search.tex]{forall-zap}
    \end{agdalisting} \columnbreak
    \begin{agdalisting}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Search.tex]{exists-zap}
    \end{agdalisting}
  \end{multicols}
\end{minipage}

This automation procedure allows us to state the property succinctly, and have
the type checker go and run the decision procedure to solve it for us.
Here's an example of its use:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{pre-inst-proof}
\end{agdalisting} \todo{Include counterexample stuff?}

% As a quick aside, when the property is \emph{not} true, for instance:
% \ExecuteMetaData[agda/Data/Pauli.tex]{contra}
% Agda will fail by not finding an instance of \(\bot\).
% The error message specifically is:
% \begin{displayquote}
%   \verb+No instance of type ⊥ was found in scope.+
% \end{displayquote}
% We can actually display a counterexample, by defining a custom empty type
% parameterised by the counterexample itself.
% % \ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{counterexample-def}
% Combined with some other changes to the combinators, this will give the
% following more helpful error message:
% \begin{displayquote}
%   \verb+No instance of type Counterexample X was found in scope.+
% \end{displayquote}
\subsection{Instances}
One bit of cruft in the above proof is the need to specify the particular
finiteness proof for bools.
While this isn't any great burden in this case, it of course becomes more
difficult in more complex circumstances.

In Agda, as in Haskell, the problem of overlapping instances often presents
itself when more complex typeclass-based machinery is used.
While there are ways to deal with overlapping instances, it is usually better to
avoid them altogether, which is precisely what we do in our approach.
We provide instances for the semiring-based combinators we defined in
\todo[noline]{reference here}
, and require the user to provide an instance for their own type.
We don't, for example, provide an instance that itself searches for surjections
from other instances: this instance is not uniquely determined.

After all of that, the change to our \(\AgdaFunction{∀↯}\) is simple: we just
change the first parameter from explicit to an instance.
% \ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{auto-inst}

The change to the \(\AgdaDatatype{Pauli}\) proof of cancellation isn't
groundbreaking at first:
\ExecuteMetaData[agda/Data/Pauli.tex]{cancel-auto}
But the real benefit is now we can automate proofs over \emph{tuples}, allowing
us to prove, for instance, commutativity.
\ExecuteMetaData[agda/Data/Pauli.tex]{comm-uncurried}
\subsection{Generic Currying and Uncurrying}
While we have arguably removed the bulk of the boilerplate from the automated
proofs, there is still the case of the ugly noise of currying and uncurrying.
In this section, we take inspiration from
\cite{allaisGenericLevelPolymorphic2019} to develop a small interface to generic
\(n\)-ary functions and properties.
What we provide here differs from that work in the following ways:
\begin{itemize}
  \item Our generic representation can handle dependent \(\sum\) and \(\prod\)
    types (rather than their non-dependent counterparts, \(\times\) and
    \(\rightarrow\)).
    This extension was necessary for our use case: it is mentioned in the paper
    as the obvious next step. 
  \item We implement the curry-uncurry combinators as (verified) isomorphisms.
    Since we are in a cubical setting, this gives us equivalences between the
    types, a feature not available in standard Agda.
  \item We deal with implicit and instance arguments generically.
\end{itemize}
A full explanation of our implementation is beyond the scope of this work, but
we will mention the key parts here.
First, we define a function arrow generic over the application method:
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{fun-arr}
Then, we prove that it is isomorphic to the normal function arrow:
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{fun-iso}
This step will allow us to write the curry-uncurry proofs once, and then extend
them to the three different argument forms without difficulty.

We do the same with dependent function types:
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{pi-iso}

Next, we need to make our machinery for multiple arguments.
Here we will define a generic tuple, indexed by some \(\Nat\).
As observed in \cite{allaisGenericLevelPolymorphic2019}, it's important to
\emph{not} implement this as an inductively defined vector, e.g.
% \ExecuteMetaData[agda/Data/Vec.tex]{def}
As this will not give us the correct \(\eta\)-equality we need for unification.
\todo[noline]{Stick reference to iterated Vec definition}
Once we have a unification-friendly vector type, we can use it to implement our
generic (and level-polymorphic) tuples.
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{tuple}
We decided against \(\top\)-terminated tuples, as the actual contents of the
tuple type are exposed in the interface.

Finally, we can prove isomorphisms between the curried and uncurried versions of
functions:
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-iso}
And dependent functions:
% \ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-pi-iso}
With these combinators, we can implement the search functions in an
arity-generic way:
% \ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{all-nary}
And automate away our proofs:
\ExecuteMetaData[agda/Data/Pauli.tex]{prf-curried}
